Pushdown Control-Flow Analysis via Refunctionalization

### 2018-02-24 paper.tex
update

Tranditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use a direct-style intermediate representation as target for
clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain $\lambda$-calculus languages.

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language in
administrative normal form (ANF) \cite{flanagan1993essence}.

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
CESK machine models program execution as state transitions in small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

It is worth noting that the continuation class \texttt{Kont} is defined as a list of frames.
The frame class \texttt{Frame} stores the information of call-site, i.e., the information that
can be used to resume the interrupted computation.
A \texttt{Frame} constitutes of a variable name \texttt{x} to be bind later, a control expression
that resumes to, and its environment.

@@

The first case is that the control of current state is a \texttt{Let} expression,
and its right-hand side is an function application.
By calling function \texttt{atomicEval}, we obtain the closure that callee \texttt{f} stands for.
The successor state's control then transfers to the \texttt{body} expression of closure,
with an updated environment and an update store. The new environment is extended
from closure's environment and mapped \texttt{v} to a fresh address \texttt{addr}.
The new store also extends with \texttt{addr} mapping to the value of \texttt{ae},
which is evaluated from \texttt{atomicEval}.
In the meanwhile, a new frame \texttt{frame} is pushed onto the stack \texttt{k}.
This \texttt{frame} contains the left-hand side variable name of \texttt{Let},
the body expression, and the lexical environment of body expression.

@@

To run the program, we use function \texttt{inject} to construct an initial machine
state given an expression. Then the \texttt{drive} function is used to evaluate
to a final state by iteratively apply \texttt{step} on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine is a systematic methodology that derives abstract interpreters
from a concrete abstract machine. TODO

@@

def atomicEval(e: Expr, env: Env, bstore: BStore): Set[Storable] = e match {
  case Var(x) => bstore(env(x)).toSet
  case lam@Lam(x, body) => Set(Clos(lam, env))
}
\end{verbatim}

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, env_c, newBStore, new_time)                                                          
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, newBStore, newKStore, newKAddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) if isAtomic(ae)=>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, new_time)
        val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as program is abstractly executing.
The following code shows a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen} set
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

@@

\subsection{Call/Return Match in Control-Flow Analysis}

describe problem, show some examples, why it is important

@@

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = f_env + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}

\end{verbatim}

@@

In another direction, we can transform the refunctionalized AAM to direct style, but explictly use 
side effects such as assignment and mutation to handle collecting semantics and non-determinism.

@@

>>>>

Tranditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.

To begin, we now present the concrete syntax of a call-by-value $\lambda$-calculus language 
in ANF \cite{flanagan1993essence}.

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
CESK machine models program execution as state transitions in a small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

It is worth noting that the continuation class \texttt{Kont} is defined as a list of frames.
The frame class \texttt{Frame} stores the information of call-site, i.e., the information that
can be used to resume the interrupted computation.
A \texttt{Frame} constitutes of a variable name \texttt{x} to be bind later, a control expression
that resumes to, and its environment. (TODO: mention that frame list vs KArg/KFun)

@@

The first case is that the control of current state is a \texttt{Let} expression,
and its right-hand side is an function application.
By calling function \texttt{atomicEval}, we obtain the closure that callee \texttt{f} stands for.
The successor state's control then transfers to the \texttt{body} expression of closure,
with an updated environment and an update store. The new environment is extended
from closure's environment and mapped \texttt{v} to a fresh address \texttt{addr}.
The new store also extends with \texttt{addr} mapping to the value of \texttt{ae},
which is evaluated from \texttt{atomicEval}.
In the meanwhile, a new frame \texttt{frame} is pushed onto the stack \texttt{k}.
This \texttt{frame} contains the variable name \texttt{x} at left-hand side position of \texttt{Let},
the body expression of \texttt{Let}, and the lexical environment of the body expression.

@@

To run the program, we use function \texttt{inject} to construct an initial machine
state given an expression \texttt{e}.
Then the \texttt{drive} function is used to evaluate
to a final state by iteratively apply \texttt{step} on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine is a systematic methodology that derives sound abstract interpreters
from a concrete abstract machine \cite{van2012systematic, van2010abstracting}.

@@

def atomicEval(e: Expr, env: Env, bstore: BStore): Set[Storable] = e match {
  case Var(x) => bstore(env(x))
  case lam@Lam(x, body) => Set(Clos(lam, env))
}
\end{verbatim}

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val newBStore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val newKAddr = allocKont(body, env_c, newBStore, new_time)
        val newKStore = kstore.update(newKAddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, newBStore, newKStore, newKAddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      for (Cont(Frame(x, e, f_env), f_kaddr) <- kstore(kaddr).toList) yield {
        val baddr = allocBind(x, new_time)
        val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen} set
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

@@

\subsection{Pushdown Control-Flow Analysis}

Pushdown control-flow is a property in analysis that precisely models
the runtime call-stack of the analyzed program. 
A pushdown control-flow analysis provides as exact as runtime return-flow 
when analyzing the program, but traditional control-flow analysis collapse 
the state space using finite space abstraction thus cause imprecise stack modeling.

To see the problem, we can consider the following example:

\begin{verbatim}
(let ([x (id 1)])
  (let ([y (id 2)])
    x))
\end{verbatim}

return flow of \texttt{id} is polluted, and returns to both \texttt{x} and
\texttt{y}. TODO: explain more

@@

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, f_env)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = f_env + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{verbatim}

@@

We can transform the refunctionalized AAM to direct style, but explictly use 
side effects such as assignment and mutation to handle collecting semantics and non-determinism.

@@

\subsection{Store Widening}

\subsection{Higher-Order Controls?}

TODO: analyze programs with call/cc, shift, reset.
